Decoupled wood and leaf strategies among canopy and understorey tree in a low-elevation rainforest of India

### 2018-07-09 draft.Rmd
Uppangala 0.1.9047 Bulletproofing code & ToDo.md

```{r Traits}
genPFT() %>% 
  select(-Tree, -Sp, -SpCode, -CE, -SHA, -LTD) %>% 
  melt(variable.name = 'Abbreviation') %>% 
      left_join(data.frame(
    Trait = c('Leaf thickness', 'Leaf area', 'Leaf dry matter content', 'Specific leaf area', 'Wood density'),
    Abbreviation = c('LT', 'LA', 'LDMC', 'SLA', 'WD'),
    Unit = c('$\\mu m$', '$mm^2$', '$mg.g^-1$', '$m^2.kg^-1$', '$g.cm^-3$'),
    Strategy = c('Leaf defense and investment',
                 'Leaf resource capture vs.investment',
                 'Leaf defense and investment',
                 'Leaf resource acquisiton vs. defense',
                 'Stem transport, structure and defense')
  )) %>% 
  group_by(Trait, Abbreviation, Unit, Strategy) %>% 
  summarise_if(is.numeric, funs(min(.,na.rm = T), max(., na.rm = T))) %>% 
  mutate(range = paste(round(min,2), "-", round(max,2))) %>% 
  select(-min, -max) %>% 
  left_join(genPFT() %>% 
              select(-Sp, -SpCode, -CE, -SHA, -LTD) %>% 
              melt(id.vars = 'Tree', 
                   variable.name = 'Abbreviation', 
                   value.name = 'Intra-individuals') %>% 
              group_by(Tree, Abbreviation) %>% 
              summarise_each(funs(sd(., na.rm = T)/mean(., na.rm = T))) %>% 
              ungroup() %>% 
              select(-Tree) %>% 
              group_by(Abbreviation) %>% 
              summarise_each(funs(mean(., na.rm = T)))) %>% 
  left_join(genPFT() %>% 
              group_by(Tree, Sp, SpCode, CE) %>% 
              summarise_each(funs(mean)) %>% 
              ungroup() %>% 
              select(-Tree, -SpCode, -CE, -SHA, -LTD) %>% 
              melt(id.vars = 'Sp', 
                   variable.name = 'Abbreviation', value.name = 'Intra-species') %>% 
              group_by(Sp, Abbreviation) %>% 
              summarise_each(funs(sd(., na.rm = T)/mean(., na.rm = T))) %>% 
              ungroup() %>% 
              select(-Sp) %>% 
              group_by(Abbreviation) %>% 
              summarise_each(funs(mean(., na.rm = T)))) %>% 
  left_join(genPFT() %>% 
              group_by(Tree, Sp, SpCode, CE) %>% 
              summarise_each(funs(mean)) %>% 
              ungroup() %>% 
              select(-Tree, -CE) %>% 
              group_by(Sp, SpCode) %>% 
              summarise_each(funs(mean)) %>% 
              ungroup() %>% 
              select(-Sp, -SpCode, -SHA, -LTD) %>% 
              melt(variable.name = 'Abbreviation', value.name = 'Inter-species') %>% 
              group_by(Abbreviation) %>% 
              summarise_each(funs(sd(., na.rm = T)/mean(., na.rm = T)))) %>%
  mutate(`Intra-individuals` = ifelse(Abbreviation == "WD", NA, `Intra-individuals`)) %>% 
  kable(digits = 2, format.args = list(big.mark = " "), format = 'pandoc',
        caption = 'Functional traits measured. Traits with their abbreviation, standard unit, associated species strategy, range of values, intra-individual, intra-specific and inter-specific component of variation.  Species strategy are precised for functioning, survival and reproduction [@baraloto_decoupled_2010]. The rlast columns represents the coefficient of variation of trait values among leaves per individuals (intra-individual), among trees per species (intra-specific), and among trees of different species.')

```

```{r Q}
# Sp level PFT
PFT <- genPFT() %>% 
  mutate(LA = log(LA)) %>% 
  mutate(SLA = log(SLA)) %>% 
  group_by(Tree, Sp, SpCode, CE) %>% 
  summarise_each(funs(mean(., na.rm = T))) %>% 
  ungroup() %>% 
  group_by(Sp, SpCode) %>% 
  summarise_each(funs(mean(., na.rm = T))) %>% 
  ungroup() %>% 
  select(-Tree, -Sp, -CE, -LTD, -SHA)
row.names(PFT) <- PFT$SpCode
# PCA
PFT.pca <- princomp(~ LA + WD + LT + LDMC + SLA, data = PFT, cor = T)
# Matrix Q
Q <- PFT %>% 
  left_join(data.frame(PFT.pca$scores) %>% mutate(SpCode = row.names(.))) %>% 
  rename(WES = Comp.1, LES = Comp.2) %>% 
  left_join(genSpecies()) %>% 
  mutate(strata = ifelse(Strata == 3, "U", "C")) %>%
  select(SLA, LDMC, WD, LA, LT, LES, WES, strata, SpCode) %>%
  mutate(strata = ifelse(SpCode == "agin", "U", strata))
```

```{r L}
com <- quadrats(cs = c(30,30))
XY <- genTrees()[c('x', 'y')]
coordinates(XY) <- ~ x + y
proj4string(XY) <- crs(com)
L <- genTrees() %>% 
  mutate(com = (XY %over% com)[,1]) %>% 
  filter(!is.na(com)) %>% 
  group_by(com, SpCode) %>% 
  summarise(n = n()) %>% 
  dcast(com ~ SpCode) %>% 
  replace(is.na(.), 0)
```

@@

```{r GWMboxplot, fig.cap="Boxplot of guilds weighted means (GWM) among communties for each traits (for canopy (C) and understorey (U) guilds). See table \\@ref(tab:Traits) for abbreviations.", fig.height=4}
L %>%
  melt(id.var = "com",
       variable.name = "SpCode",
       value.name = "abundance") %>% 
  dcast(SpCode ~ com) %>% 
  inner_join(Q) %>%
  select(-SpCode) %>%
  melt(id.vars = c('strata', 'LES', 'SLA', 'WES', 'WD', 'LDMC', 'LA', 'LT'),
       variable.name = 'community', value.name = 'abundance') %>%
  group_by(strata, community) %>%
  summarise_each(funs(weighted.mean(., abundance, na.rm = T)), -abundance) %>% 
  select(-community, -LES, -WES) %>% 
  melt(id.vars = "strata", 
       variable.name = "trait",
       value.name = "GWM") %>% 
  ggplot(aes(x = strata, y = GWM, fill = strata)) +
  geom_boxplot() +
  facet_wrap(~trait, scales = "free", nrow = 1) +
  xlab("Guild") + scale_fill_discrete(guide = "none") +
  ggpubr::stat_compare_means(aes(label = paste(..p.signif..))) 
```

@@

```{r GWMses}
GWM <- L %>%
  melt(id.var = "com",
       variable.name = "SpCode",
       value.name = "abundance") %>% 
  dcast(SpCode ~ com) %>% 
  left_join(Q) %>% 
  select(-SpCode) %>% 
  melt(id.vars = c('strata', 'LES', 'SLA', 'WES', 'WD', 'LDMC', 'LA', 'LT'),
       variable.name = 'community', value.name = 'abundance') %>% 
  group_by(strata, community) %>% 
  summarise_each(funs(weighted.mean(., abundance, na.rm = T)), -abundance)
# nQ <- array(rep(as.matrix(Q), n),
#             c(dim(Q), n),
#             list(row.names(Q), names(Q), 1:n))
# nQ <- array(apply(nQ, 3, function(x){
#   x[which(x[,"strata"] == "C"),'SpCode'] <- sample(x[which(x[,"strata"] == "C"),'SpCode'])
#   x[which(x[,"strata"] == "U"),'SpCode'] <- sample(x[which(x[,"strata"] == "U"),'SpCode'])
#   return(x)
#   }), c(dim(Q), n), list(row.names(Q), names(Q), 1:n))
# cl <- makeCluster(cores)
# clusterExport(cl, list('L', 'nQ'))
# nGWM <- parApply(cl, nQ, 3, function(x){
#   library(dplyr)
#   library(reshape2)
#   L %>%
#     melt(id.var = "com",
#          variable.name = "SpCode",
#          value.name = "abundance") %>%
#     dcast(SpCode ~ com) %>%
#     left_join(data.frame(x)) %>%
#     select(-SpCode) %>%
#     melt(id.vars = c('strata', 'LES', 'SLA', 'WES', 'WD', 'LDMC', 'LA', 'LT'),
#          variable.name = 'community', value.name = 'abundance') %>%
#     group_by(strata, community) %>%
#     mutate_each(funs(as.character), -abundance) %>%
#     mutate_each(funs(as.numeric), -abundance) %>%
#     summarise_each(funs(weighted.mean(., abundance, na.rm = T)), -abundance)
# })
# stopCluster(cl)
# nGWM <- array(unlist(lapply(nGWM, function(x) as.matrix(x[3:9]))),
#               c(dim(GWM[3:9]), n),
#               list(row.names(GWM), names(GWM[3:9]), 1:n))
# rm(cl, nQ)
# save(nGWM, file = file.path("draft_save", 'nGWM.Rdata'))
load(file.path("draft_save", 'nGWM.Rdata'))
GWM_ses <- GWM
GWM_ses[3:9] <- (GWM[3:9] - apply(nGWM, 1:2, mean, na.rm = T))/apply(nGWM, 1:2, sd, na.rm = T)
```

```{r LM}
GWM_ses %>% 
  filter(!is.na(strata)) %>% 
  melt(id.vars = c('strata', 'community')) %>% 
  rename(trait = variable) %>% 
  rename(SES = value) %>% 
  left_join(R) %>%
  group_by(strata, trait) %>% 
  do(reg = step(lm(SES ~ Slope + Curvature + Wetness + SW + Canopy, .), 
                     direction = 'both', trace = 0)) %>% 
  tidy(reg) %>% 
  group_by(strata, trait, term) %>%
  transmute(value = paste(format(estimate, scientific = T, digits = 2, trim = T), 
                          stars(p.value, ns = '   '))) %>% 
  dcast(strata + trait ~ term) %>% 
  select(strata, trait, Slope, Curvature, Wetness, SW, Canopy) %>% 
  mutate_each(funs(replace(., is.na(.), ' '))) %>% 
  left_join( # Inclusion of R2 but should be split
    GWM_ses %>% 
      filter(!is.na(strata)) %>% 
      melt(id.vars = c('strata', 'community')) %>% 
      rename(trait = variable) %>% 
      rename(SES = value) %>% 
      left_join(R) %>%
      group_by(strata, trait) %>% 
      do(reg = step(lm(SES ~ Slope + Curvature + Wetness + SW + Canopy, .), 
                    direction = 'both', trace = 0)) %>% 
      glance(reg) %>% 
      select(r.squared) %>% 
      mutate(r.squared = round(r.squared, 2)) %>% 
      rename("$R^2$" = r.squared)
  )%>% # End of R2
  filter(trait != 'LES') %>% 
  filter(trait != 'WES') %>%
  bold() %>% 
  kable(caption = 'Linear models of the standardized effect size (SES) with null model NM2 for each trait of GWMs, for canopy (C) and understorey (U) guilds, in abundance. We performed stepwise selection of the predictors including slope, curvature, wetness southwesterness (SW), and canopy height (Canopy), and only the selected predictors are shown. See table \\@ref(tab:Traits) for abbreviations. Bold and number of asterisk indicates the significance threshold ( * p-value<0.05, ** p-value <0.01, *** p-value<0.001).', format = 'pandoc')
```

>>>>

```{r Traits}
genPFT() %>% 
  mutate(id = row.names(.)) %>% 
  select(-Tree, -Sp, -SpCode, -CE, -SHA, -LTD) %>% 
  melt(id.vars = "id", variable.name = 'Abbreviation') %>% 
  mutate(Abbreviation = as.character(Abbreviation)) %>% 
  left_join(data.frame(
    Trait = c('Leaf thickness', 'Leaf area', 'Leaf dry matter content', 'Specific leaf area', 'Wood density'),
    Abbreviation = c('LT', 'LA', 'LDMC', 'SLA', 'WD'),
    Unit = c('$\\mu m$', '$mm^2$', '$mg.g^-1$', '$m^2.kg^-1$', '$g.cm^-3$'),
    Strategy = c('Leaf defense and investment',
                 'Leaf resource capture vs.investment',
                 'Leaf defense and investment',
                 'Leaf resource acquisiton vs. defense',
                 'Stem transport, structure and defense'),
    stringsAsFactors = F
  ),  by = "Abbreviation") %>% 
  group_by(Trait, Abbreviation, Unit, Strategy) %>% 
  summarise_if(is.numeric, funs(min(.,na.rm = T), max(., na.rm = T))) %>% 
  mutate(range = paste(round(min,2), "-", round(max,2))) %>% 
  select(-min, -max) %>% 
  left_join(genPFT() %>% 
              select(-Sp, -SpCode, -CE, -SHA, -LTD) %>% 
              melt(id.vars = 'Tree', 
                   variable.name = 'Abbreviation', 
                   value.name = 'Intra-individuals') %>% 
              mutate(Abbreviation = as.character(Abbreviation)) %>% 
              group_by(Tree, Abbreviation) %>% 
              summarise_all(funs(sd(., na.rm = T)/mean(., na.rm = T))) %>% 
              ungroup() %>% 
              select(-Tree) %>% 
              group_by(Abbreviation) %>% 
              summarise_all(funs(mean(., na.rm = T))),  by = "Abbreviation") %>% 
  left_join(genPFT() %>% 
              group_by(Tree, Sp, SpCode, CE) %>% 
              summarise_all(funs(mean)) %>% 
              ungroup() %>% 
              select(-Tree, -SpCode, -CE, -SHA, -LTD) %>% 
              melt(id.vars = 'Sp', 
                   variable.name = 'Abbreviation', value.name = 'Intra-species') %>% 
              mutate(Abbreviation = as.character(Abbreviation)) %>% 
              group_by(Sp, Abbreviation) %>% 
              summarise_all(funs(sd(., na.rm = T)/mean(., na.rm = T))) %>% 
              ungroup() %>% 
              select(-Sp) %>% 
              group_by(Abbreviation) %>% 
              summarise_all(funs(mean(., na.rm = T))),  by = "Abbreviation") %>% 
  left_join(genPFT() %>% 
              group_by(Tree, Sp, SpCode, CE) %>% 
              summarise_all(funs(mean)) %>% 
              ungroup() %>% 
              select(-Tree, -CE) %>% 
              group_by(Sp, SpCode) %>% 
              summarise_all(funs(mean)) %>% 
              ungroup() %>% 
              select(-Sp, -SpCode, -SHA, -LTD) %>%
              mutate(id = row.names(.)) %>% 
              melt(id.vars = "id", 
                   variable.name = 'Abbreviation', value.name = 'Inter-species') %>% 
              select(-id) %>% 
              mutate(Abbreviation = as.character(Abbreviation)) %>% 
              group_by(Abbreviation) %>%
              summarise_all(funs(sd(., na.rm = T)/mean(., na.rm = T))),
            by = "Abbreviation") %>%
  mutate(`Intra-individuals` = ifelse(Abbreviation == "WD", NA, `Intra-individuals`)) %>% 
  kable(digits = 2, format.args = list(big.mark = " "), format = 'pandoc',
        caption = 'Functional traits measured. Traits with their abbreviation, standard unit, associated species strategy, range of values, intra-individual, intra-specific and inter-specific component of variation.  Species strategy are precised for functioning, survival and reproduction [@baraloto_decoupled_2010]. The rlast columns represents the coefficient of variation of trait values among leaves per individuals (intra-individual), among trees per species (intra-specific), and among trees of different species.')
```

```{r Q}
# Sp level PFT
PFT <- genPFT() %>% 
  mutate(LA = log(LA)) %>% 
  mutate(SLA = log(SLA)) %>% 
  group_by(Tree, Sp, SpCode, CE) %>% 
  summarise_all(funs(mean(., na.rm = T))) %>% 
  ungroup() %>% 
  select(-Tree, -CE) %>% 
  group_by(Sp, SpCode) %>% 
  summarise_all(funs(mean(., na.rm = T))) %>% 
  ungroup() %>% 
  select(-Sp, -LTD, -SHA)
suppressWarnings(row.names(PFT) <- PFT$SpCode) # deprecated
# PCA
PFT.pca <- princomp(~ LA + WD + LT + LDMC + SLA, data = PFT, cor = T)
# Matrix Q
Q <- PFT %>% 
  mutate(SpCode = as.character(SpCode)) %>% 
  left_join(data.frame(PFT.pca$scores) %>% 
              mutate(SpCode = row.names(.)),
            by = "SpCode") %>% 
  rename(WES = Comp.1, LES = Comp.2) %>% 
  left_join(genSpecies(), by = "SpCode") %>% 
  mutate(strata = ifelse(Strata == 3, "U", "C")) %>%
  select(SLA, LDMC, WD, LA, LT, LES, WES, strata, SpCode) %>%
  mutate(strata = ifelse(SpCode == "agin", "U", strata))
```

```{r L}
com <- quadrats(cs = c(30,30))
XY <- genTrees()[c('x', 'y')]
coordinates(XY) <- ~ x + y
proj4string(XY) <- crs(com)
L <- genTrees() %>% 
  mutate(com = (XY %over% com)[,1]) %>% 
  filter(!is.na(com)) %>% 
  group_by(com, SpCode) %>% 
  summarise(n = n()) %>% 
  dcast(com ~ SpCode, value.var = "n") %>% 
  replace(is.na(.), 0)
```

@@

```{r GWMboxplot, fig.cap="Boxplot of guilds weighted means (GWM) among communties for each traits (for canopy (C) and understorey (U) guilds). See table \\@ref(tab:Traits) for abbreviations.", fig.height=4}
L %>%
  melt(id.var = "com",
       variable.name = "SpCode",
       value.name = "abundance") %>% 
  dcast(SpCode ~ com, value.var = "abundance") %>% 
  mutate(SpCode = as.character(SpCode)) %>% 
  inner_join(Q, by = "SpCode") %>%
  select(-SpCode) %>%
  melt(id.vars = c('strata', 'LES', 'SLA', 'WES', 'WD', 'LDMC', 'LA', 'LT'),
       variable.name = 'community', value.name = 'abundance') %>%
  group_by(strata, community) %>%
  summarise_each(funs(weighted.mean(., abundance, na.rm = T)), -abundance) %>% 
  select(-community, -LES, -WES) %>% 
  melt(id.vars = "strata", 
       variable.name = "trait",
       value.name = "GWM") %>% 
  ggplot(aes(x = strata, y = GWM, fill = strata)) +
  geom_boxplot() +
  facet_wrap(~trait, scales = "free", nrow = 1) +
  xlab("Guild") + scale_fill_discrete(guide = "none") +
  ggpubr::stat_compare_means(aes(label = paste(..p.signif..))) 
```

@@

```{r GWMses}
GWM <- L %>%
  melt(id.var = "com",
       variable.name = "SpCode",
       value.name = "abundance") %>% 
  dcast(SpCode ~ com, value.var = "abundance") %>% 
  mutate(SpCode = as.character(SpCode)) %>% 
  left_join(Q, by = "SpCode") %>% 
  select(-SpCode) %>% 
  melt(id.vars = c('strata', 'LES', 'SLA', 'WES', 'WD', 'LDMC', 'LA', 'LT'),
       variable.name = 'community', value.name = 'abundance') %>% 
  group_by(strata, community) %>% 
  summarise_each(funs(weighted.mean(., abundance, na.rm = T)), -abundance)
# nQ <- array(rep(as.matrix(Q), n),
#             c(dim(Q), n),
#             list(row.names(Q), names(Q), 1:n))
# nQ <- array(apply(nQ, 3, function(x){
#   x[which(x[,"strata"] == "C"),'SpCode'] <- sample(x[which(x[,"strata"] == "C"),'SpCode'])
#   x[which(x[,"strata"] == "U"),'SpCode'] <- sample(x[which(x[,"strata"] == "U"),'SpCode'])
#   return(x)
#   }), c(dim(Q), n), list(row.names(Q), names(Q), 1:n))
# cl <- makeCluster(cores)
# clusterExport(cl, list('L', 'nQ'))
# nGWM <- parApply(cl, nQ, 3, function(x){
#   library(dplyr)
#   library(reshape2)
#   L %>%
#     melt(id.var = "com",
#          variable.name = "SpCode",
#          value.name = "abundance") %>%
#     dcast(SpCode ~ com) %>%
#     left_join(data.frame(x)) %>%
#     select(-SpCode) %>%
#     melt(id.vars = c('strata', 'LES', 'SLA', 'WES', 'WD', 'LDMC', 'LA', 'LT'),
#          variable.name = 'community', value.name = 'abundance') %>%
#     group_by(strata, community) %>%
#     mutate_each(funs(as.character), -abundance) %>%
#     mutate_each(funs(as.numeric), -abundance) %>%
#     summarise_each(funs(weighted.mean(., abundance, na.rm = T)), -abundance)
# })
# stopCluster(cl)
# nGWM <- array(unlist(lapply(nGWM, function(x) as.matrix(x[3:9]))),
#               c(dim(GWM[3:9]), n),
#               list(row.names(GWM), names(GWM[3:9]), 1:n))
# rm(cl, nQ)
# save(nGWM, file = file.path("draft_save", 'nGWM.Rdata'))
load(file.path("draft_save", 'nGWM.Rdata'))
GWM_ses <- GWM
GWM_ses[3:9] <- (GWM[3:9] - apply(nGWM, 1:2, mean, na.rm = T))/apply(nGWM, 1:2, sd, na.rm = T)
```

```{r LM}
GWM_ses %>% 
  filter(!is.na(strata)) %>% 
  melt(id.vars = c('strata', 'community')) %>% 
  rename(trait = variable) %>% 
  rename(SES = value) %>% 
  mutate(community = as.character(community)) %>% 
  left_join(R, by = "community") %>%
  group_by(strata, trait) %>% 
  do(reg = step(lm(SES ~ Slope + Curvature + Wetness + SW + Canopy, .), 
                     direction = 'both', trace = 0)) %>% 
  tidy(reg) %>% 
  group_by(strata, trait, term) %>%
  transmute(value = paste(format(estimate, scientific = T, digits = 2, trim = T), 
                          stars(p.value, ns = '   '))) %>% 
  dcast(strata + trait ~ term) %>% 
  select(strata, trait, Slope, Curvature, Wetness, SW, Canopy) %>% 
  mutate_all(funs(replace(., is.na(.), ' '))) %>% 
  left_join( # Inclusion of R2 but should be split
    GWM_ses %>% 
      filter(!is.na(strata)) %>% 
      melt(id.vars = c('strata', 'community')) %>% 
      rename(trait = variable) %>% 
      rename(SES = value) %>% 
      mutate(communtiy = as.character(community)) %>% 
      left_join(R, by = "community") %>%
      group_by(strata, trait) %>% 
      do(reg = step(lm(SES ~ Slope + Curvature + Wetness + SW + Canopy, .), 
                    direction = 'both', trace = 0)) %>% 
      glance(reg) %>% 
      select(r.squared) %>% 
      mutate(r.squared = round(r.squared, 2)) %>% 
      rename("$R^2$" = r.squared), by = c("strata", "trait")
  )%>% # End of R2
  filter(trait != 'LES') %>% 
  filter(trait != 'WES') %>%
  bold() %>% 
  kable(caption = 'Linear models of the standardized effect size (SES) with null model NM2 for each trait of GWMs, for canopy (C) and understorey (U) guilds, in abundance. We performed stepwise selection of the predictors including slope, curvature, wetness southwesterness (SW), and canopy height (Canopy), and only the selected predictors are shown. See table \\@ref(tab:Traits) for abbreviations. Bold and number of asterisk indicates the significance threshold ( * p-value<0.05, ** p-value <0.01, *** p-value<0.001).', format = 'pandoc')
```
